Performance Evaluation of Big Data Processing Strategies for Neuroimaging

### 2018-12-13 paper.tex
Made several edits to the manuscript for grammar and language.  Overall looks pretty good.

\begin{abstract}
We benchmark Big Data processing strategies, namely in-memory 
computing, data locality, and lazy evaluation, on typical neuroimaging 
processing use cases represented by the BigBrain dataset. We use Dell 
EMC's Top-500 cluster, as well as Apache Spark and Nipype as 
representative processing engines. We model Big Data thresholds by 
comparing the data/compute ratio of the application to the disk and 
shared file system bandwidths reported to the number of concurrent 
processes in a node and across the cluster. This model acknowledges the 
fact that page caching provided by the Linux kernel is critical to the 
performance of Big Data applications. Results show that in-memory 
computing alone speeds-up executions by a factor of up to 1.6, while 
combined with data locality this factor reaches 5.3. Such important 
speed-up values are likely to be observed on typical image processing 
operations performed on images of size larger than 75GB. A ballpark 
speculation from our model showed that in-memory computing alone would 
not speed-up current 
functional MRI analyses, but that it would when coupled with data 
locality, for a number of concurrently  processed subjects in the order 
of 200. In addition, we observe that emulating in-memory computing 
using in-memory file systems (tmpfs) does not reach the performance of 
an in-memory engine, presumably due to swapping to disk and the lack of data cleanup.
We conclude that Big Data processing strategies are 
worth developing for neuroimaging 
applications. 
\end{abstract}

@@

Big Data processing engines reduce data-related overheads, through 
use of concepts such as data locality, in-memory computing, and lazy evaluation.
Data locality-aware scheduling, popularized by the MapReduce~\cite{dean2008mapreduce} 
framework, runs tasks to the nodes holding the data rather than 
performing costly data transfers across the network. In-memory 
computing, adopted by MapReduce's successor Apache 
Spark~\cite{zaharia2016apache}, stores data in memory whenever 
possible, reducing costly disk I/Os between pipeline steps. 
Lazy evaluation, also used in Spark, means that results are computed only 
when necessary, enabling further performance optimizations. Frameworks 
such as
MapReduce and Spark have become mainstream tools for data analytics, 
although many others, such as 
Dask~\cite{rocklin2015dask}, are emerging. 
Meanwhile, several scientific 
domains including bioinformatics, physics or astronomy, have entered 
the Big Data era due to increasing data volumes and variety. 
However, the adoption of Big Data engines for scientific data analysis 
remains limited, perhaps due to the widespread availability of 
scientific processing engines such as Pegasus~\cite{deelman2005pegasus} or
Taverna~\cite{oinn2004taverna}, and the adaptations required in Big 
Data processing engines for scientific computing.

Scientific applications differ from the typical Big Data use 
cases, which might explain the remaining gap between Big Data and 
scientific engines. While Big Data applications mostly target text 
processing (e.g. Web search, frequent pattern mining, recommender 
systems~\cite{leskovec2014mining}) implemented in consistent software 
libraries, scientific applications often involve 
binary data such as images and signals, processed by command-line tools 
using a mix of programming languages (C, Fortran, Python, shell 
scripts). Infrastructure-wise, Big Data applications commonly run on 
clouds or dedicated commodity clusters with locality-aware file systems 
such as the Hadoop Distributed File System 
(HDFS~\cite{shvachko2010hadoop}), whereas scientific applications are 
usually deployed on large, shared clusters where data is moved between
data nodes and compute nodes using shared file systems such 
as Lustre~\cite{schwan2003lustre}. Such differences in applications and 
infrastructures have important consequences. To 
mention only one, in-memory computing requires instrumentation to be 
applied to command-line tools.

@@

Neuroimaging, our primary field of interest, is no exception to the 
generalized rise of data volumes in science, due to the joint increase 
of image resolution and subject cohort sizes~\cite{van2014human}. 
Processing engines have been developed with neuroinformatics 
applications in mind, for instance Nipype~\cite{gorgolewski2011nipype} 
or the Pipeline System for Octave and Matlab 
(PSOM~\cite{bellec2012pipeline}). Big Data engines have been used for 
neuroimaging applications too, including in the Thunder 
project~\cite{freeman2014mapping} and in more specific works such 
as~\cite{makkie2019fast}. However, no quantitative performance 
evaluation has been conducted on neuroimaging applications to assess the 
added-value of Big Data engines compared to traditional processing engines.

@@

Answers to these questions have important implications. 
In~\cite{mehta2017comparative}, a comparative study of Dask, Spark, 
TensorFlow, MyriaDB, and SciDB on neuroinformatics use-cases is 
presented. It concludes that these systems need to be extended to 
better address scientific code integration, data partitioning, data 
formats and system tuning. We argue 
that such efforts should only be conducted if substantial performance 
improvements are expected from in-memory computing, lazy 
evaluation or data 
locality. On the other hand, neuroimaging data processing engines are 
still being developed, and it is legitimate to wonder whether these 
projects should just be migrated to Spark, Dask, or other Big Data 
engines.

Our study focuses on performance. We intentionally do not intend to 
compare Big Data and scientific data processing engines on the grounds 
of workflow language expressivity, fault-tolerance, provenance capture 
and representation, portability or reproducibility, which are otherwise 
critical concerns, addressed for instance in~\cite{samba}. Besides, our 
study of performance focuses on the impact of data writes and 
transfers. It purposely leaves out task scheduling to computing 
resources, to focus on the understanding of 
data writes and movement. Task scheduling will be part of our 
discussion though.

Infrastructure-wise, we focus on the case of High-Performance Computing 
(HPC) clusters that are typically available through University 
facilities or national computing infrastructures such as 
\href{xsede.org}{XSEDE}, \href{http://computecanada.ca}{Compute Canada} 
or \href{http://www.prace-ri.eu}{PRACE}, as neuroscientists typically use such platforms.
 We assume that HPC systems are 
multi-tenant, that compute nodes are accessible through a batch 
scheduler, and that a file system shared among the compute nodes is 
available. We intentionally leave distributed, 
HDFS-like file system out, as initiatives to deploy them in HPC 
centers, for instance Hadoop on-demand~\cite{krishnan2011myhadoop}, have not 
become mainstream yet.

@@

Our methods, including performance models, processing engines, 
applications and infrastructure used are described in 
Section~\ref{sec:methods}. Section~\ref{sec:results} presents our 
results which we discuss in Section~\ref{sec:discussion} along with the 
two research questions mentioned previously. 
Section~\ref{sec:conclusion} concludes on the relevance of Big Data 
processing strategies for neuroimaging applications.

@@

Two types of operations can be performed on RDDs:
transformations and actions. Applying a transformation to an RDD produces a new,
child, RDD through a narrow or wide dependency. A narrow dependency signifies 
that the child is only dependent on a single parent partition, whereas a child 
RDD is dependent on all parent partitions in a wide dependency. Examples of 
Spark transformations include map, filter and join. To materialize an RDD, an
action must be performed, such as a reduce or a collect. \todo{Lazy evaluation should be mentioned here, or elsewhere in the paragraph.
It is currently not described.}

@@

Three different types of schedulers are compatible with Spark: 1) Spark 
Standalone, 2) YARN~\cite{vavilapalli2013apache} and 3) Mesos~\cite{hindman2011mesos}. The Spark Standalone
scheduler is the default scheduler. YARN was a scheduler designed for Hadoop~\cite{white2012hadoop}
clusters. It is prepackaged with Hadoop installations. In contrast to YARN, 
Mesos was designed to be used in multitenant cluster environments. In our experiments,
we focus on the Spark Standalone cluster.

@@

Jobs, or Interfaces, in Nipype, are encapsulated by a Node. A Node dictates that
the job will execute on a single input. However, the MapNode, 
a child variant of the Node, can execute on multiple inputs. All tasks in
Nipype execute in their own uniquely named subdirectory which facilitates the provenance 
tracking of inputs and outputs. This also enables checkpointing of the workflow.
In the case of Node failure or application modification, only the nodes who have
been modified (verified by hash) are re-executed.

@@

 The main difference between storing the data in the engine memory, as in Spark, 
 and simply writing to an in-memory file system, such as tmpfs, is 
 related to what happens when the processed data fills up the available 
 memory. In case engine memory is used, the engine must cleanup 
 unused data with some strategy to avoid crashes. In case an in-memory 
 file system is used, the kernel starts swapping memory to disk, and 
 the performance becomes that of local disk writes. In our experiments, 
 we will explore the region  when the data consumed by the 
 application approaches the threshold of available memory.

@@

This model applies to parallel applications assuming that (1) 
concurrent processes all write the same amount of data, (2) 
concurrent processes all consume the same CPU time, (3) data is written 
uniformly along task execution. With these assumptions, all the 
processes will write at the same rate, which explains why the model 
does not depend on the total number of concurrent processes in the 
application, but only on the max number of concurrent processes 
executing on the same node ($\gamma$). We realize that these 
assumptions would usually be violated in practice, but this simple 
model already provides interesting insights on the performance of disk 
writes, as shown later. Naturally, the model also ignores other 
processes that might be writing to disk concurrently to the 
application, which we assume negligible here.

In general, an application should ensure that $\dot d$ remains negative 
or null, leading to the following inequality:
\begin{equation}
\frac{D}{C} \leq \frac{\delta}{\gamma} \label{eq:page-cache-inequality}
\end{equation}
This defines a data-compute ratio beyond which the page cache becomes 
asymptotically useless. It should be noted that the transient phase 
during the page cache fills up might last a significant amount of time, 
in particular when $\dot d$ is positive but small. We intentionally do not model the 
transient phase as it requires detailed knowledge on the size of the 
page cache and the initial amount 
of data in it ($d_0$), which are difficult to estimate.

@@

 All experiments were executed on  
 \href{https://www.dellemc.com/resources/en-us/asset/sales-documents/products/storage/h16221-hpc-lab-brochure.pdf}{Dell EMC's Zenith 
 cluster}, a Top-500 machine in the Dell EMC HPC Innovation Lab, running 
 SLURM. For the Spark experiments, a Spark cluster was started atop our 
 SLURM allocation. We had access to 16 dedicated nodes. Each compute 
 node ran on RHEL 7.4, was 
equipped with 40 Intel Xeon Gold 6148 2.4GHz CPUs and had
187GB of memory, 94GB of tmpfs and a Dell 120GB M.2 SATA SSD as local disk. 
A 698TB Lustre server was accessible on each node through a 100Gb/s 
Infiniband network. The apparent write bandwidth of the local disk, 
Lustre file system and tmpfs were measured by sequentially writing various numbers
of image blocks containing random intensities, to avoid 
caching effects (measure\_bandwidth.py script in folder 
benchmark\_scripts). They are reported in Table~\ref{table:bdwdths}.

@@

% BigBrain
% Some BIDS dataset for fmriprep
We used BigBrain~\cite{amunts2013bigbrain}, a 75GB 40$\mu$m isotropic 
histological image of a 65-year-old male's brain. The BigBrain was selected due 
to being the only one of its kind, as there does not yet exist a 
higher-resolution image of a human brain. Moreover, there currently exists a 
lack of standardized tools for processing the BigBrain as a consequence 
of its size. To examine the effects processing BigBrain has on the Big Data 
strategies, we partitioned the full $3845\times3015\times3470$ voxel image into 30 ($5\times3\times2$)
chunks, 125 ($5\times5\times5$) chunks and 750 ($5\times15\times10$) chunks. Additionally, the full 
image was also split in half ($769\times603\times347$ voxels) and the half image was 
partitioned into 125 chunks.

>>>>

\begin{abstract}
We benchmark in-memory, computing, data locality and lazy Big Data processing 
strategies on typical neuroimaging processing use cases represented by the BigBrain dataset. We use Dell 
EMC's Top-500 cluster, as well as Apache Spark and Nipype as 
representative processing engines. We model Big Data thresholds by 
comparing the data/compute ratio of the application to the disk and 
shared file system bandwidths reported to the number of concurrent 
processes in a node and across the cluster. This model acknowledges the 
fact that page caching provided by the Linux kernel is critical to the 
performance of Big Data applications. Results show that in-memory 
computing alone speeds-up executions by a factor of up to 1.6, while 
combined with data locality this factor reaches 5.3. Such important 
speed-up values are likely to be observed on typical image processing 
operations performed on images of size larger than 75GB. A ballpark 
speculation from our model showed that in-memory computing alone would 
not speed-up current 
functional MRI analyses, but that it would when coupled with data 
locality, for a number of concurrently  processed subjects in the order 
of 200. In addition, we observe that emulating in-memory computing 
using in-memory file systems (tmpfs) does not reach the performance of 
an in-memory engine, presumably due to swapping to disk and the lack of data cleanup.
We conclude that Big Data processing strategies are 
worth developing for neuroimaging 
applications. 
\end{abstract}

@@

Big Data processing engines reduce data-related overheads through 
use of concepts such as data locality, in-memory computing, and lazy evaluation.
Data locality-aware scheduling, popularized by the MapReduce~\cite{dean2008mapreduce} 
framework, runs tasks to the nodes holding the data rather than 
performing costly data transfers across the network. In-memory 
computing, adopted by MapReduce's successor Apache 
Spark~\cite{zaharia2016apache}, stores data in memory whenever 
possible, reducing costly disk I/Os between pipeline steps. 
Lazy evaluation, also used in Spark, means that results are computed only 
when necessary to enable further performance optimizations. Frameworks 
such as
MapReduce and Spark have become mainstream tools for data analytics, 
although many others, such as 
Dask~\cite{rocklin2015dask}, are emerging. 
Meanwhile, several scientific 
domains including bioinformatics, physics or astronomy, have entered 
the Big Data era due to increasing data volumes and variety. 
However, the adoption of Big Data engines for scientific data analysis 
remains limited, perhaps due to the widespread availability of 
scientific processing engines such as Pegasus~\cite{deelman2005pegasus} or
Taverna~\cite{oinn2004taverna}, and the adaptations required in Big 
Data processing engines for scientific computing.

Scientific applications differ from the typical Big Data use 
cases, which might explain the remaining gap between Big Data and 
scientific engines. While Big Data applications mostly target text 
processing (e.g. web search, frequent pattern mining, recommender 
systems~\cite{leskovec2014mining}) implemented in consistent software 
libraries, scientific applications often involve 
binary data such as images and signals, processed by command-line tools 
using a mix of programming languages (C, Fortran, Python, shell 
scripts). Infrastructure-wise, Big Data applications commonly run on 
clouds or dedicated commodity clusters with locality-aware file systems 
such as the Hadoop Distributed File System 
(HDFS~\cite{shvachko2010hadoop}), whereas scientific applications are 
usually deployed on large, shared clusters where data is moved between
data nodes and compute nodes using shared file systems such 
as Lustre~\cite{schwan2003lustre}. Such differences in applications and 
infrastructures have important consequences. To 
mention only one, in-memory computing requires instrumentation to be 
applied to command-line tools.

@@

Our primary field of interest, Neuroimaging, is no exception to the 
generalized rise of data volumes in science due to the joint increase 
of image resolution and subject cohort sizes~\cite{van2014human}. 
Processing engines have been developed with neuroinformatics 
applications in mind, for instance Nipype~\cite{gorgolewski2011nipype} 
or the Pipeline System for Octave and Matlab 
(PSOM~\cite{bellec2012pipeline}). Big Data engines have been used for 
neuroimaging applications too, including in the Thunder 
project~\cite{freeman2014mapping} and in more specific works such 
as~\cite{makkie2019fast}. However, no quantitative performance 
evaluation has been conducted on neuroimaging applications to assess the 
added-value of Big Data engines compared to traditional processing engines.

@@

Answers to these questions have important implications. 
In~\cite{mehta2017comparative}, a comparative study of Dask, Spark, 
TensorFlow, MyriaDB, and SciDB on neuroinformatics use-cases is 
presented. It concludes that these systems need to be extended to 
better address scientific code integration, data partitioning, data 
formats and system tuning. We argue 
that such efforts should only be conducted if substantial performance 
improvements are expected from in-memory computing, lazy 
evaluation or data 
locality. On the other hand, neuroimaging data processing engines are 
still being developed, and the question remains whether these 
projects should just be migrated to Spark, Dask, or other Big Data 
engines.

Our study focuses on performance. We intentionally do not 
compare Big Data and scientific data processing engines on the grounds 
of workflow language expressivity, fault-tolerance, provenance capture 
and representation, portability or reproducibility which are otherwise 
critical concerns, addressed for instance in~\cite{samba}. Besides, our 
study of performance focuses on the impact of data writes and 
transfers. It purposely leaves out task scheduling to computing 
resources, to focus on the understanding of 
data writes and movement. Task scheduling will be part of our 
discussion though.

Infrastructure-wise, we focus on the case of High-Performance Computing 
(HPC) clusters that are typically available through University 
facilities or national computing infrastructures such as 
\href{xsede.org}{XSEDE}, \href{http://computecanada.ca}{Compute Canada} 
or \href{http://www.prace-ri.eu}{PRACE}, as neuroscientists typically use such platforms.
 We assume that HPC systems are 
multi-tenant, that compute nodes are accessible through a batch 
scheduler, and that a file system shared among the compute nodes is 
available. We intentionally did not consider distributed, 
HDFS-like file system, as initiatives to deploy them in HPC 
centers, for instance Hadoop on-demand~\cite{krishnan2011myhadoop}, have not 
become mainstream yet.

@@

Our methods, including performance models, processing engines, 
applications and infrastructure used, are described in 
Section~\ref{sec:methods}. Section~\ref{sec:results} presents our 
results which we discuss in Section~\ref{sec:discussion} along with the 
two research questions mentioned previously. 
Section~\ref{sec:conclusion} concludes on the relevance of Big Data 
processing strategies for neuroimaging applications.

@@

Two types of operations can be performed on RDDs:
transformations and actions. Applying a transformation to an RDD produces a new
child RDD through a narrow or wide dependency. A narrow dependency signifies 
that the child is only dependent on a single parent partition, whereas a child 
RDD is dependent on all parent partitions in a wide dependency. Examples of 
Spark transformations include map, filter and join. To materialize an RDD, an
action must be performed, such as a reduce or a collect. \todo{Lazy evaluation should be mentioned here, or elsewhere in the paragraph.
It is currently not described.}

@@

Three different types of schedulers are compatible with Spark: 1) Spark 
Standalone, 2) YARN~\cite{vavilapalli2013apache} and 3) Mesos~\cite{hindman2011mesos}. The Spark Standalone
scheduler is the default scheduler. YARN was a scheduler designed for Hadoop~\cite{white2012hadoop}
clusters and is prepackaged with Hadoop installations. In contrast to YARN, 
Mesos was designed to be used in multitenant cluster environments. In our experiments,
we focus on the Spark Standalone cluster.

@@

Jobs, or Interfaces, in Nipype, are encapsulated by a Node. A Node dictates that
the job will execute on a single input. However, the MapNode, 
a child variant of the Node, can execute on multiple inputs. All tasks in
Nipype execute in their own uniquely named subdirectory which facilitates provenance 
tracking of inputs and outputs, and also enables checkpointing of the workflow.
In the case of Node failure or application modification, only the nodes who have
been modified (verified by hash) are re-executed.

@@

 The main difference between storing the data in the engine memory, as in Spark, 
 and simply writing to an in-memory file system, such as tmpfs, is 
 what happens when the processed data fills up the available 
 memory. In case engine memory is used, the engine must cleanup 
 unused data with some strategy to avoid crashes. In case an in-memory 
 file system is used, the kernel starts swapping memory to disk, and 
 the performance becomes that of local disk writes. In our experiments, 
 we will explore the region  when the data consumed by the 
 application approaches the threshold of available memory.

@@

This model applies to parallel applications assuming that (1) 
concurrent processes all write the same amount of data, (2) 
concurrent processes all consume the same CPU time, (3) data is written 
uniformly along task execution. With these assumptions, all the 
processes will write at the same rate, which explains why the model 
does not depend on the total number of concurrent processes in the 
application, but only on the max number of concurrent processes 
executing on the same node ($\gamma$). While these 
assumptions would usually be violated in practice, this simple 
model already provides interesting insights on the performance of disk 
writes, as shown later. Naturally, the model also ignores other 
processes that might be writing to disk concurrently to the 
application, which we assume negligible here.

In general, an application should ensure that $\dot d$ remains negative 
or null, leading to the following inequality:
\begin{equation}
\frac{D}{C} \leq \frac{\delta}{\gamma} \label{eq:page-cache-inequality}
\end{equation}
This defines a data-compute ratio beyond which the page cache becomes 
asymptotically useless. It should be noted that the transient phase 
during which the page cache fills up might last a significant amount of time, 
in particular when $\dot d$ is positive and small. We intentionally do not model the 
transient phase as it requires detailed knowledge of difficult to estimate parameters  
such as the page cache and the initial amount 
of data in it ($d_0$).

@@

 All experiments were executed on  
 \href{https://www.dellemc.com/resources/en-us/asset/sales-documents/products/storage/h16221-hpc-lab-brochure.pdf}{Dell EMC's Zenith 
 cluster}, a Top-500 machine in the Dell EMC HPC Innovation Lab, running 
 SLURM. For the Spark experiments, a Spark cluster was started atop our 
 SLURM allocation through access to 16 dedicated nodes. Each compute 
 node ran on RHEL 7.4, was 
 equipped with 40 Intel Xeon Gold 6148 2.4GHz CPUs and had
 187GB of memory, 94GB of tmpfs and a Dell 120GB M.2 SATA SSD as local disk. 
 A 698TB Lustre server was accessible on each node through a 100Gb/s 
 Infiniband network. The apparent write bandwidth of the local disk, 
 Lustre file system and tmpfs were measured by sequentially writing various numbers
 of image blocks containing random intensities, to avoid 
 caching effects (measure\_bandwidth.py script in folder 
 benchmark\_scripts). They are reported in Table~\ref{table:bdwdths}.

@@

% BigBrain
% Some BIDS dataset for fmriprep
We used BigBrain~\cite{amunts2013bigbrain}, a 75GB 40$\mu$m isotropic 
histological image of a 65-year-old male's brain. The BigBrain was selected due 
to being the only one of its kind, as there does not yet exist any other
higher-resolution image of a human brain. Moreover, there currently exists a 
lack of standardized tools for processing the BigBrain as a consequence 
of its size. To examine the effects processing the BigBrain has on Big Data 
strategies, we partitioned the full $3845\times3015\times3470$ voxel image into 30 ($5\times3\times2$)
chunks, 125 ($5\times5\times5$) chunks and 750 ($5\times15\times10$) chunks. Additionally, the full 
image was also split in half ($769\times603\times347$ voxels) and the half image was 
partitioned into 125 chunks.
